\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-18 rebuttal.tex

}
\opening{Dear chair and members of the reading committee}
\todo{fix todos in document}

I would like to extend my sincere gratitude for time and effort you have taken
to review the manuscript of my PhD dissertation.
Additionally, I would like to thank you for the insightful, constructive and
detailed comments you all have provided.
The following reply contains a point-by-point response to your review comments.
Attached, you will also find an updated draft of the manuscript and a version
with indicated changes for your approval.

\reply{Andrea K\"ubler}
Major comments:
\begin{enumerate}
	\item
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item
\end{enumerate}

\reply{Fabien Lotte}
Major comments:
\begin{enumerate}
	\item Part of the confusion between active and passive BCI might be due
	to the poor choice of phrasing ``paradigms that require little active
  participation from the user, like redirecting attention or initiating imagined actions, are classified as passive BCI.''.
	This has been rewritten to ``in contrast to paradigms that require the user's active participation, for
	instance through redirecting attention or initiating imagined actions,
	paradigms free of this requirement are classified as passive bci,''
	for clarity.
	Neurofeedback has been reclassified as active BCI in the text and in
	figure 1.3.
	\item \todo{Why XDAWN+RG no shrinkage (performed worse), fair
	comparison of sample rate}
  \item Appendix 1.A now fully reports all obtained accuracies to give a better
    insight in effect sizes. Wherever the significant effects have been mentioned
    in the text, the effect size has now been reported along with it.
	\item In the results section of chapter 4, statistical comparisons were
	only performed between methods for which we evaluated the performance
	ourselves (HODA, PARAFACDA and BTTDA).
	Since the MOABB benchmarking framework~\cite{Aristimunha2023} allows for
	fair comparison of performance metrics reported across different works,
	the analyses for other comparison methods were not re-run
	by us as reported, but performances were
	taken from the MOABB benchmark database~\cite{Chevallier2024}.
	However, this source provided only the aggregated performance metrics
	and not the metrics per subject, session and cross-validation fold,
	hence we could not perform relevant statistical comparison with these
	comparison models.
	\item \todo{include table with absolute performances for transfer
	conditions}
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item According to the annotations, typographical errors have been
	corrected in text and figures, and imprecise or inaccurate statements have been adjusted.
	The recommended references have been discussed and, where necessary,
	statements were updated accordingly.
  Figure 1.5 was corrected and matched with the correct caption.
\end{enumerate}

\reply{Reinhold Scherer}
Major comments:
\begin{enumerate}
	\item \todo{quantify qualitative statements}
	\item  Mathematical equations and derivations have thoroughly
	been checked for correctness and notational consistency, and have been
	corrected throughout the manuscript where necessary.
	\item As indicated by the reviewer, true on-line experiments with live
	feedback have indeed not yet be conducted, but were foreseen to take
	place within the framework of this doctoral study.
	I agree that these experiments would be valuable to supplement the
	presented conclusions and are the logical next step.
	Indeed, the start of experimentation with healthy control participants
	as well as patient groups, was heavily delayed by the COVID-19 pandemic, as the
	preparatory research started in September 2019 and I officially
	enrolled in the the doctoral training programme in January 2020.
	This restricted not only our access to experiment subjects for a
	significant period at the start of the PhD, and even longer for
	vulnerable patient groups, but also shifted priorities for the medical
	staff at our partner institutions and patient centers.
	Hence, work early in the PhD mainly focused on the development of
	computational methods.
	We are currently starting up true on-line experimentation in cooperation
	with our partner patient centers, with an adapted interface and experimental protocol.
  This limitation has now clearly been mentioned in the future work section of
  chapter 8.
	\item Recorded datasets could not be made publicly available since the
	ethical protocol (Ethics Commission of University Hospital Leuven --
	S62547 v1.4) for the conducted studies did not include provisions
	publishing the data.
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item
\end{enumerate}

@@

\reply{Adalberto Simeone}
Major comments:
\begin{enumerate}
	\item \todo{discuss in limitations the impact of letters and future
	work experiments with an actual usable interface}
	\todo{mention here plans to go to patients with actual speller
	interface}
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item Measurements unit representation has been uniformized according to the
    SI standard.
  \item Numerical citations have now consistently been replaced by text-style citations
    mentioning the author names where appropriate in the sentence.
  \item Contractions have been remove throughout the manuscript.
\end{enumerate}

### 2024-11-18 chapters/bci.tex
Textcites

People with \ac{sspi} due to neurodegenerative disease, stroke, traumatic brain or spine
injury become entirely reliant on caregivers or family members in their day-to-day
routine.
Wouldn't it be great if we could help them regain some quality of life by making them
more independent?

@@

The key problem for severely paralyzed individuals is that the mind wants to go where the
body cannot follow.
Couldn't we, therefore, design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of technology, like
a robot arm or speech synthesizer, interacting with the world on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also termed brain-machine interface
(BMI) when coupled to a physical actuator, like a robot arm or an exoskeleton.
The term \ac{bci} is usually preferred for assistive technology and communication
devices.}, a system that directly couples actions in the `real world' to the mental state
of the user.

@@

The \ac{bci} paradigms mentioned above can also be combined to gain performance.
Straightforward examples are activating or deactivating \ac{ssvep} stimulation with a
motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
along multiple dimensions at once.
\textcite{Han2023} recently used this strategy, combining frequency and phase coding in
\ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an efficient \ac{bci}
where one of 200 targets can be accurately selected with only 800ms of stimulation.

@@

Band-pass filtering is then usually applied to isolate the frequency ranges of interest
by removing frequencies that are too high or too low to contain relevant neural signals.
Band-pass filters eliminate slow drifts and high-frequency noise, such as muscle
artifacts or environmental electrical interference, improving the \ac{snr}.
Additionally, notch filtering can be used to remove strong power line interference\footnote{%
Typically at 50Hz in Europe.}.
This interference can heavily contaminate \ac{eeg} recordings and even leak through when
outside the passband of a previous filter.
Notch filtering efficiently attenuates this narrowband noise without affecting the
surrounding frequencies.

@@

The visual oddball paradigm is a \emph{reactive}, \emph{stimulus-dependent},
\ac{bci} paradigm with all the benefits and drawbacks.
Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
The brain response of interest is the visual \ac{erp}, which can be accurately measured
and decoded from non-invasive \ac{eeg} signals with a relatively low computational
effort and short calibration time.
First established by~\cite{Farwell1988} in 1988, it is well supported by literature and
has been proven to work for those with \acl{sspi} in day-to-day home use~\cite{Wolpaw2018}.

@@

One by one, visual elements on a computer screen are \emph{intensified} for a short
period of time by changing in color, brightness, or size.
Intensifications of different targets are usually 100-200ms apart and last for about
100ms~\cite{Sellers2006a}.
On each flash observed by the user, specific brain activity is elicited.
If one of these intensified elements is rare with respect to the others, i.e., it is the
\emph{odd-one out} or \emph{oddball}, this brain activity is altered.
A stimulus can be rare either due to its inherent properties like color or location.
Crucially, this is also the case if it is `marked' as rare by the user, e.g., by paying
explicit attention to only a given stimulus and not the others or counting how many
times it occurs.
If these visual elements represent letters, and we know the timing of the
intensifications of each letter, we can now establish communication by detecting for
which letter an oddball response was present in the brain signal at its time of
intensification.

\Ac{itr} in the oddball paradigm is optimized by intensifying multiple targets at once
in a row-column strategy and using a sequence of selections, giving rise to the classic
matrix speller of~\cite{Farwell1988}.
Other optimizations increase response \ac{snr}, like using flashing face images as
intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
stimuli~\cite{Treder2011}.
The number of targets in a visual oddball \ac{bci} is limited by the crowding effect,
which imposes a limit on how close targets can be to each other while not distracting
the user's attention~\cite{Sellers2006a,Li2010}.
This may be overcome by making hierarchical selections of stimuli representing groups
of selections~\cite{Treder2010}.

@@

The time-domain response elicited immediately after the intensification of a stimulus is
known as the visual \ac{erp}~\cite{Luck2014}.
An \ac{erp} is a waveform consisting of multiple components.
Some of these components are modulated by whether the stimulus was an oddball or not.
These components appear as positive peaks and negative troughs in the \ac{erp} waveform.
They follow a naming scheme based on polarity (Negative or Positive) and latency (e.g.,
N2 or N200 for a negative component after $\pm200$ ms).
The component most prominently modulated by attention is called the P3 (positive
deflection after $\pm300$ ms).
Some \ac{bci}\footnote{In \ac{erp} analysis, components are sometimes referred to with
the timing around which they occur (e.g., N170, negative after 170ms), or by their rank
in order of occurrence (N1, the first negative component).
The timing nomenclature is based on average latencies for neurotypical individuals, but
is seldom correct for specific \ac{bci} users due to the large variability across
\ac{erp} stimulation paradigms and subjects, and can also depend on the user's pathology.
Furthermore, as we will see later, this work focuses specifically on the intra-session
variability in \ac{erp} latency, explicitly assuming the latencies are not fixed.
Therefore, we adhere to the ranking nomenclature.
Fortunately, for the visual oddball paradigm, the rankings do roughly correspond to
their expected timings, i.e., P1=P100, N1=N100, etc.}.
Therefore, oddball paradigms are frequently referred to as P3(00) paradigms.
However, due to the equally important contributions of other \ac{erp} components in
decoding, as we will show later, this could be considered a misnomer.
Thus, we will adhere to the oddball paradigm naming.

@@

\subsection{Preprocessing \& feature extraction}
For \ac{erp} analysis, the \ac{eeg} is usually band-pass filtered between 0.5Hz and
16Hz.
Re-referencing can be done to average mastoids to highlight the contribution of the P3
component.
In addition to the steps outlined in \cref{sec:bci/preprocessing}, the signal is
usually cut into \emph{epochs} for \ac{erp} analysis.
These are time windows surrounding the stimulus event.
They usually include some baseline interval before the stimulus, which can be used to
normalize the response.

>>>>

People with \ac{sspi} due to neurodegenerative disease, stroke, traumatic brain or spine
injury become entirely reliant on caregivers or family members in their day-to-day
routine.
Would it not be great if we could help them regain some quality of life by making them
more independent?

@@

The key problem for severely paralyzed individuals is that the mind wants to go where the
body cannot follow.
Could we therefore not design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of technology, like
a robot arm or speech synthesizer, interacting with the world on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also termed brain-machine interface
(BMI) when coupled to a physical actuator, like a robot arm or an exoskeleton.
The term \ac{bci} is usually preferred for assistive technology and communication
devices.}, a system that directly couples actions in the `real world' to the mental state
of the user.

@@

The \ac{bci} paradigms mentioned above can also be combined to gain performance.
Straightforward examples are activating or deactivating \ac{ssvep} stimulation with a
motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
along multiple dimensions at once.
\textcite{Han2023} recently used this strategy, combining frequency and phase coding in
\ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an efficient \ac{bci}
where one of 200 targets can be accurately selected with only 800 ms of stimulation.

@@

Band-pass filtering is then usually applied to isolate the frequency ranges of interest
by removing frequencies that are too high or too low to contain relevant neural signals.
Band-pass filters eliminate slow drifts and high-frequency noise, such as muscle
artifacts or environmental electrical interference, improving the \ac{snr}.
Additionally, notch filtering can be used to remove strong power line interference\footnote{%
Typically at 50 Hz in Europe.}.
This interference can heavily contaminate \ac{eeg} recordings and even leak through when
outside the passband of a previous filter.
Notch filtering efficiently attenuates this narrowband noise without affecting the
surrounding frequencies.

@@

The visual oddball paradigm is a \emph{reactive}, \emph{stimulus-dependent},
\ac{bci} paradigm with all the benefits and drawbacks.
Nevertheless, it can score relatively high on the \ac{itr} parameters established above.
The brain response of interest is the visual \ac{erp}, which can be accurately measured
and decoded from non-invasive \ac{eeg} signals with a relatively low computational
effort and short calibration time.
First established by \textcite{Farwell1988} in 1988, it is well supported by literature and
has been proven to work for those with \acl{sspi} in day-to-day home use~\cite{Wolpaw2018}.

@@

One by one, visual elements on a computer screen are \emph{intensified} for a short
period of time by changing in color, brightness, or size.
Intensifications of different targets are usually 100-200 ms apart and last for about
100 ms~\cite{Sellers2006a}.
On each flash observed by the user, specific brain activity is elicited.
If one of these intensified elements is rare with respect to the others, i.e., it is the
\emph{odd-one out} or \emph{oddball}, this brain activity is altered.
A stimulus can be rare either due to its inherent properties like color or location.
Crucially, this is also the case if it is `marked' as rare by the user, e.g., by paying
explicit attention to only a given stimulus and not the others or counting how many
times it occurs.
If these visual elements represent letters, and we know the timing of the
intensifications of each letter, we can now establish communication by detecting for
which letter an oddball response was present in the brain signal at its time of
intensification.

\Ac{itr} in the oddball paradigm is optimized by intensifying multiple targets at once
in a row-column strategy and using a sequence of selections, giving rise to the classic
matrix speller of \textcite{Farwell1988}.
Other optimizations increase response \ac{snr}, like using flashing face images as
intensifications~\cite{Jin2012} or adding distinct colors and shapes to the
stimuli~\cite{Treder2011}.
The number of targets in a visual oddball \ac{bci} is limited by the crowding effect,
which imposes a limit on how close targets can be to each other while not distracting
the user's attention~\cite{Sellers2006a,Li2010}.
This may be overcome by making hierarchical selections of stimuli representing groups
of selections~\cite{Treder2010}.

@@

The time-domain response elicited immediately after the intensification of a stimulus is
known as the visual \ac{erp}~\cite{Luck2014}.
An \ac{erp} is a waveform consisting of multiple components.
Some of these components are modulated by whether the stimulus was an oddball or not.
These components appear as positive peaks and negative troughs in the \ac{erp} waveform.
They follow a naming scheme based on polarity (Negative or Positive) and latency (e.g.,
N2 or N200 for a negative component after $\pm200$ ms).
The component most prominently modulated by attention is called the P3 (positive
deflection after $\pm300$ ms).
Some \ac{bci}\footnote{In \ac{erp} analysis, components are sometimes referred to with
the timing around which they occur (e.g., N170, negative after 170 ms), or by their rank
in order of occurrence (N1, the first negative component).
The timing nomenclature is based on average latencies for neurotypical individuals, but
is seldom correct for specific \ac{bci} users due to the large variability across
\ac{erp} stimulation paradigms and subjects, and can also depend on the user's pathology.
Furthermore, as we will see later, this work focuses specifically on the intra-session
variability in \ac{erp} latency, explicitly assuming the latencies are not fixed.
Therefore, we adhere to the ranking nomenclature.
Fortunately, for the visual oddball paradigm, the rankings do roughly correspond to
their expected timings, i.e., P1=P100, N1=N100, etc.}.
Therefore, oddball paradigms are frequently referred to as P3(00) paradigms.
However, due to the equally important contributions of other \ac{erp} components in
decoding, as we will show later, this could be considered a misnomer.
Thus, we will adhere to the oddball paradigm naming.

@@

\subsection{Preprocessing \& feature extraction}
For \ac{erp} analysis, the \ac{eeg} is usually band-pass filtered between 0.5 Hz and
16 Hz.
Re-referencing can be done to average mastoids to highlight the contribution of the P3
component.
In addition to the steps outlined in \cref{sec:bci/preprocessing}, the signal is
usually cut into \emph{epochs} for \ac{erp} analysis.
These are time windows surrounding the stimulus event.
They usually include some baseline interval before the stimulus, which can be used to
normalize the response.

### 2024-11-18 chapters/bttda.tex
Textcites

\section{Experiments}
\subsection{Datasets and decoders}
We evaluated our proposed model in two offline EEG-based BCI decoding problems:
the event-related potential (ERP) and motor imagery (MI) paradigms using a
selection of the openly available MOABB benchmarking datasets~\cite{Aristimunha2023}.
Two ERP and two MI datasets were retained to reduce computational
demand.
Details about these datasets are found in \cref{tab:moabb}.
For the ERP datasets, the task is to distinguish target from non-target ERPs,
while the MI datasets consist of distinguishing different imagined limb
movements.
Within-session classification performance was assessed using stratified 5-fold
cross-validation to calculate the area under the receiver operating
characteristic curve (ROC-AUC).
\begin{table*}[t]
	\centering
	\footnotesize
  \makebox[\textwidth][c]{
	  \input{chapters/bttda/moabb.tex}
  }
  \caption[MOABB datasets used for evaluation.]{MOABB datasets used for evaluation, with the number of
		subjects (\# Sub.), the number of EEG channels (\# Chan.), the number of
    classes (\# Classes), the number of trials or trials per class for ERP
    datasets (\# Trials), the epoch length (Epoch len.), the sampling
		frequency (S. freq.), the number of sessions per subject (\# Sess.) and the
		number of runs (\# Runs).
    Adapted from~\cite{Aristimunha2023}
    and~\cite{Chevallier2024}.}
	\label{tab:moabb}
\end{table*}

@@

Furthermore, as additional comparison methods, we used the methods evaluated in~\cite{Chevallier2024}.
For the ERP datasets, these were the Riemannian Geometry-based methods
using augmented ERP covariance matrices with and without SVD dimensionality
reduction features for a Minimum Riemannian Distance to Mean classifier
(ERPCov+MDM and ERPCovSVD+MDM), augmented ERP covariance matrices after
XDAWN~\cite{Rivet2009}
filtering paired with a Riemannian Minimum Distance to Mean classifier or a
projection to tangent space and a support vector machine
(XDAWNCov+MDM and XDAWNCov+TS+SVM), and LDA applied after XDAWN filtering
(XDAWN+LDA).
For the MI datasets, the comparison methods were selected from Riemannian
methods.
These include projection onto the Riemannian tangent space combined with an
ElasticNet classifier (TS+EL), a Fisher Geodesic Minimum Riemannian Distance to
Mean classifier and the classification of augmented covariance matrices
projected onto the Riemannian tangent space with a support vector machine
(ACM+TS+SVM).
Additionally, we report the performance of the Common Spatial Patterns filter
and a frequency filterbank combined with a support vector machine (CSP+SVM and
FilterBank+SVM, respectively).
We refer to~\cite{Chevallier2024} for implementation details of these comparison
methods.

\subsection{Event-Related Potentials}
ERPs are spatiotemporal features, with each sample forming a $2^\text{nd}$
order tensor with $K=2$ modes (a matrix), representing EEG channels and time samples
per epoch.
EEG signals for the evaluated datasets were recorded at the sample rate given
by \cref{tab:moabb} and band-pass filtered between 1Hz
and 24Hz.
The signals were cut into epochs starting from stimulus onset with a
dataset-specific length given by \cref{tab:moabb}.
For HODA+LDA, PARAFACDA+LDA, and BTTDA+LDA decoders, epochs were downsampled to 48Hz.
For the BNCI2014-008 and BNCI2015-003 datasets, this resulted in matrices of
dimensionality $(8,48)$ and $(8,38)$, respectively.

\Cref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
decoders.
\begin{table}[t]
	\footnotesize
	\centering
	\input{chapters/bttda/score_erp.tex}
  \caption[Within-session classification score for 2 event-related potential
  datasets.]{Area under the receiver operating characteristic curve for
		cross-validated within-session evaluation for HODA and our proposed decoders
		PARAFACDA and BTTDA evaluated on 2 event-related potential datasets.
    Scores for other decoders were taken from \cite{Chevallier2024}.
		BTTDA reaches the highest performance for both evaluated datasets, closely
		followed by PARAFACDA.
	}
	\label{tab:erp-score}
\end{table}
The highest performance is achieved with the proposed BTTDA+LDA.
One-sided Wilcoxon signed-rank tests with a significance level of $\alpha=0.05$
reveal that both PARAFACDA+LDA and BTTDA+LDA significantly outperform
HODA+LDA in both the BNCI2014-008 (PARAFACDA+LDA: $p=0.0039$, BTTDA+LDA:
$p=0.0039$) and the BNCI2015-003 (PARAFACDA+LDA: $p=0.0001$, BTTDA+LDA:
$p=0.0049$) datasets.
No significant difference was found between BTTDA+LDA and PARAFACDA+LDA.

@@

The MI datasets listed in \cref{tab:moabb} were first band-pass filtered
between 8 and 30Hz and cut into
epochs with time windows as specified by \cref{tab:moabb}.
Next, time-frequency transformation was performed using a complex Morlet wavelet
convolution, with 16 wavelet frequencies logarithmically spaced between 8 and
32Hz.
The number of wavelet cycles $c$ varied with wavelet frequency $f$ as
$c=0.7*f$.
Features were extracted by taking the log-transformed envelope of the complex
wavelet transformation and averaging each epoch along the time axis into time bins of
length $1/4$s.
For the BNCI2014-001 and BNCI2014-004 datasets, this resulted in tensors of dimensionality
$(22, 4, 16)$ and $(3, 4, 18)$, respectively.

\cref{tab:mi-score} lists the cross-validated classification scores for
the evaluated motor imagery datasets.
Note that, in line with~\cite{Chevallier2024}, accuracy is reported for the
multi-class classification problem in BNCI2014-001, while ROC-AUC was reported for the
binary classification problem in BNCI2014-004.
\begin{table}[t]
  \centering
	\footnotesize
	\input{chapters/bttda/score_mi.tex}
  \caption[Within-session classification score for 2 motor imagery datasets.]{Classification score for
		cross-validated within-session evaluation for HODA+LDA and our proposed decoders
		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
		Scores for other decoders were taken from \cite{Chevallier2024}. Accuracy
    is listed for BNCI2014-001, and area under the receiver-operator
    characteristic curve for BNCI2014-004. BTTDA outperforms HODA and PARAFACDA
    for BNCI2014-001 but does not reach performance comparable to the current
    state-of-the-art. For BNCI2014-004, PARAFACDA and BTTDA perform
    approximately on par with HODA and the current state-of-the art.}
	\label{tab:mi-score}
\end{table}
For the BNCI2014-001 dataset, HODA+LDA and our proposed decoders do not reach
satisfactory performance compared to the comparison methods, yet both
PARAFACDA+LDA and BTTDA+LDA improve upon HODA+LDA, and for BTTDA+LDA this
difference is significant ($p=0.0269$).
For the BNCI2014-004 dataset, the performance gap with comparison methods is
smaller, but no significant differences were found between HODA+LDA,
PARAFACDA+LDA, and BTTDA+LDA\todo{qualitative}.

>>>>

\section{Experiments}
\subsection{Datasets and decoders}
We evaluated our proposed model in two offline EEG-based BCI decoding problems:
the event-related potential (ERP) and motor imagery (MI) paradigms using a
selection of the openly available MOABB benchmarking datasets~\cite{Aristimunha2023}.
Two ERP and two MI datasets were retained to reduce computational
demand.
Details about these datasets are found in \cref{tab:moabb}.
For the ERP datasets, the task is to distinguish target from non-target ERPs,
while the MI datasets consist of distinguishing different imagined limb
movements.
Within-session classification performance was assessed using stratified 5-fold
cross-validation to calculate the area under the receiver operating
characteristic curve (ROC-AUC).
\begin{table*}[t]
	\centering
	\footnotesize
  \makebox[\textwidth][c]{
	  \input{chapters/bttda/moabb.tex}
  }
  \caption[MOABB datasets used for evaluation.]{MOABB datasets used for evaluation, with the number of
		subjects (\# Sub.), the number of EEG channels (\# Chan.), the number of
    classes (\# Classes), the number of trials or trials per class for ERP
    datasets (\# Trials), the epoch length (Epoch len.), the sampling
		frequency (S. freq.), the number of sessions per subject (\# Sess.) and the
		number of runs (\# Runs).
    Adapted from \textcite{Aristimunha2023} and \textcite{Chevallier2024}.}
	\label{tab:moabb}
\end{table*}

@@

Furthermore, as additional comparison methods, we used the methods evaluated
by~\textcite{Chevallier2024}.
For the ERP datasets, these were the Riemannian Geometry-based methods
using augmented ERP covariance matrices with and without SVD dimensionality
reduction features for a Minimum Riemannian Distance to Mean classifier
(ERPCov+MDM and ERPCovSVD+MDM), augmented ERP covariance matrices after
XDAWN~\cite{Rivet2009}
filtering paired with a Riemannian Minimum Distance to Mean classifier or a
projection to tangent space and a support vector machine
(XDAWNCov+MDM and XDAWNCov+TS+SVM), and LDA applied after XDAWN filtering
(XDAWN+LDA).
For the MI datasets, the comparison methods were selected from Riemannian
methods.
These include projection onto the Riemannian tangent space combined with an
ElasticNet classifier (TS+EL), a Fisher Geodesic Minimum Riemannian Distance to
Mean classifier and the classification of augmented covariance matrices
projected onto the Riemannian tangent space with a support vector machine
(ACM+TS+SVM).
Additionally, we report the performance of the Common Spatial Patterns filter
and a frequency filterbank combined with a support vector machine (CSP+SVM and
FilterBank+SVM, respectively).
We refer to \textcite{Chevallier2024} for implementation details of these comparison
methods.

\subsection{Event-Related Potentials}
ERPs are spatiotemporal features, with each sample forming a $2^\text{nd}$
order tensor with $K=2$ modes (a matrix), representing EEG channels and time samples
per epoch.
EEG signals for the evaluated datasets were recorded at the sample rate given
by \cref{tab:moabb} and band-pass filtered between 1 Hz
and 24 Hz.
The signals were cut into epochs starting from stimulus onset with a
dataset-specific length given by \cref{tab:moabb}.
For HODA+LDA, PARAFACDA+LDA, and BTTDA+LDA decoders, epochs were downsampled to 48 Hz.
For the BNCI2014-008 and BNCI2015-003 datasets, this resulted in matrices of
dimensionality $(8,48)$ and $(8,38)$, respectively.

\Cref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
decoders.
\begin{table}[t]
	\footnotesize
	\centering
	\input{chapters/bttda/score_erp.tex}
  \caption[Within-session classification score for 2 event-related potential
  datasets.]{Area under the receiver operating characteristic curve for
		cross-validated within-session evaluation for HODA and our proposed decoders
		PARAFACDA and BTTDA evaluated on 2 event-related potential datasets.
    Scores for other decoders were taken from \textcite{Chevallier2024}.
		BTTDA reaches the highest performance for both evaluated datasets, closely
		followed by PARAFACDA.
	}
	\label{tab:erp-score}
\end{table}
The highest performance is achieved with the proposed BTTDA+LDA.
One-sided Wilcoxon signed-rank tests with a significance level of $\alpha=0.05$
reveal that both PARAFACDA+LDA and BTTDA+LDA significantly outperform
HODA+LDA in both the BNCI2014-008 (PARAFACDA+LDA: $p=0.0039$, BTTDA+LDA:
$p=0.0039$) and the BNCI2015-003 (PARAFACDA+LDA: $p=0.0001$, BTTDA+LDA:
$p=0.0049$) datasets.
No significant difference was found between BTTDA+LDA and PARAFACDA+LDA.

@@

The MI datasets listed in \cref{tab:moabb} were first band-pass filtered
between 8 and 30 Hz and cut into
epochs with time windows as specified by \cref{tab:moabb}.
Next, time-frequency transformation was performed using a complex Morlet wavelet
convolution, with 16 wavelet frequencies logarithmically spaced between 8 and
32 Hz.
The number of wavelet cycles $c$ varied with wavelet frequency $f$ as
$c=0.7*f$.
Features were extracted by taking the log-transformed envelope of the complex
wavelet transformation and averaging each epoch along the time axis into time bins of
length $1/4$s.
For the BNCI2014-001 and BNCI2014-004 datasets, this resulted in tensors of dimensionality
$(22, 4, 16)$ and $(3, 4, 18)$, respectively.

\cref{tab:mi-score} lists the cross-validated classification scores for
the evaluated motor imagery datasets.
Note that, in line with~\textcite{Chevallier2024}, accuracy is reported for the
multi-class classification problem in BNCI2014-001, while ROC-AUC was reported for the
binary classification problem in BNCI2014-004.
\begin{table}[t]
  \centering
	\footnotesize
	\input{chapters/bttda/score_mi.tex}
  \caption[Within-session classification score for 2 motor imagery datasets.]{Classification score for
		cross-validated within-session evaluation for HODA+LDA and our proposed decoders
		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
		Scores for other decoders were taken from \textcite{Chevallier2024}. Accuracy
    is listed for BNCI2014-001, and area under the receiver-operator
    characteristic curve for BNCI2014-004. BTTDA outperforms HODA and PARAFACDA
    for BNCI2014-001 but does not reach performance comparable to the current
    state-of-the-art. For BNCI2014-004, PARAFACDA and BTTDA perform
    approximately on par with HODA and the current state-of-the art.}
	\label{tab:mi-score}
\end{table}
For the BNCI2014-001 dataset, HODA+LDA and our proposed decoders do not reach
satisfactory performance compared to the comparison methods, yet both
PARAFACDA+LDA and BTTDA+LDA improve upon HODA+LDA, and for BTTDA+LDA this
difference is significant ($p=0.0269$).
For the BNCI2014-004 dataset, the performance gap with comparison methods is
smaller, but no significant differences were found between HODA+LDA,
PARAFACDA+LDA, and BTTDA+LDA\todo{qualitative}.

### 2024-11-18 chapters/conclusion.tex
Textcites

In gaze-independence decoding, our current efforts focus on,
counterintuitively, integrating eye tracking into the decoding strategy.
Since we are able to explicitly discern overt, covert, and split \ac{vsa} from
the \ac{erp}, and different decoders perform best in different settings, it
could be helpful to derive the current setting from the gaze position and the
\ac{erp}, and propose the most suited decoder for a given data point.
This could allow us to select the best classifier using eye tracking.

@@

Finally, we wish to expand our experimentation with individuals with
\ac{sspgi}.
We aim to revisit the participants with an on-line experiment and comfort
assessment to establish the usability of the proposed interface and compare its
\ac{itr} to the state-of-the-art literature.

@@

Unfortunately, this is also probably one of the least practical ways to set up
a 4-year research project, such as in the context of a doctoral thesis.
This approach can only work if the proper facilities can be gathered immediately
from the start, such as ethical approval, access to patients, and on-line \acp{bci}
experimentation systems.
It is therefore important that research labs maintain a long-term working
collaboration with patient centers, so that new projects can immediately be
verified with the user population. This also allows interested users to compare
multiple systems and guide development.
The fact that a user has previous \ac{bci} experience is often seen as a confounding
factor in off-line studies.
When the research is in a phase that heavily depends on user
experimentation, this is actually a strength, since the user can help the researcher gain
insight into what works and doesn’t work for them, and where research efforts
should be focused.

>>>>

In gaze-independence decoding, our current efforts focus on --
counterintuitively -- integrating eye tracking into the decoding strategy.
Since we are able to explicitly discern overt, covert, and split \ac{vsa} from
the \ac{erp}, and different decoders perform best in different settings, it
could be helpful to derive the current setting from the gaze position and the
\ac{erp}, and propose the most suited decoder for a given data point.
This could allow us to select the best classifier using eye tracking.

@@

Finally, we wish to expand our experimentation with individuals with
\ac{sspgi}.
As the work presented in this thesis started during the COVID-19 pandemic,
experimentation was delayed and the envisioned application could not be
validated within the time frame of the doctoral project.
To complete this work, we aim to revisit the participants with an on-line
experiment and comfort assessment to establish the usability of the proposed
interface and compare its \ac{itr} to the state-of-the-art literature.

@@

Unfortunately, this is also probably one of the least practical ways to set up
a 4-year research project, such as in the context of a doctoral thesis.
This approach can only work if the proper facilities can be gathered immediately
from the start, such as ethical approval, access to patients, and on-line \acp{bci}
experimentation systems.
It is therefore important that research labs maintain a long-term working
collaboration with patient centers, so that new projects can immediately be
verified with the user population. This also allows interested users to compare
multiple systems and guide development.
The fact that a user has previous \ac{bci} experience is often seen as a confounding
factor in off-line studies.
When the research is in a phase that heavily depends on user
experimentation, this is actually a strength, since the user can help the researcher gain
insight into what works and what does not work for them, and where research efforts
should be focused.
